---
title: "Adobe Streamflow Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adobe Streamflow Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(adobeCreekData)
library(tidyverse)
library(lubridate)
```

# Introdcuction 

Give a little intro about the the purpose of the model and the overall results.

# Data

The data being used to develop this model is USGS streamflow data. The following 
gages were used:

* Highland Creek - 11449010 (1966-1977)
* Kelsey Creek - 11449500 (1970-present)
* Adobe Creek - 11448500 (1954-1978)

The Highland Creek gage is below Highland dam, the Adobe Creek gage is above the 
Adobe Creek dam. All this data is bundled within this data package, more information 
for each is available with: 

```{r}
?adobeCreekData::kelsey_creek_flow
```

In order to calculate Adobe Creek **after** the Highland Creek junction we summed 
up the two flows. 

```{r}
adobe_creek_downstream <- 
  adobeCreekData::adobe_creek_flow %>% 
  left_join(adobeCreekData::highland_creek_flow, by=c("date" = "date"), 
            suffix = c("_adobe", "_highland")) %>% 
  select(date, flow_cfs_highland, flow_cfs_adobe) %>% 
  mutate(adobe_downstream_cfs = flow_cfs_highland + flow_cfs_adobe) %>% 
  filter(!is.na(adobe_downstream_cfs))
```

This resulted in the following streamflow

```{r}
adobe_creek_downstream %>% 
  filter(year(date) == 1975, month(date) %in% 2:8) %>% 
  ggplot() + 
  geom_line(aes(date, adobe_downstream_cfs, color="Adobe Downstream")) + 
  geom_line(aes(date, flow_cfs_highland, color="Highland Creek")) + 
  geom_line(aes(date, flow_cfs_adobe, color="Adobe Creek")) + 
  labs(x = "Date", y = "Flow (cfs)", color="")
```

# Modeling Approach

# Modeling

## Model Training

## Model Evaluation 

## Results


# Conclusion
